dyngen: a multi-modal simulator for spearheading new single-cell omics analyses

### 2021-05-17 manuscript/manuscript.tex
update

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  table,
  10pt,
  a4paper]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={dyngen: a multi-modal simulator for spearheading new single-cell omics analyses},
  pdfauthor={Robrecht Cannoodt,1,2,3; Wouter Saelens,1,2; Louise Deconinck1,2; Yvan Saeys1,2,},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{tcolorbox}
\usepackage{colortbl}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{fontspec}
\usepackage{pifont}
\usepackage{float}
\usepackage{caption}
\captionsetup[table]{position=bottom}
\setmainfont [Path = fonts/,
  UprightFont = *-300,
  ItalicFont = *-300-Italic,
  BoldFont = *-700,
  BoldItalicFont = *-700-Italic
]{MuseoSans}

\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[sorting=none,url=false,backend=bibtex]{biblatex}

\usepackage{filecontents}

\bibliography{\jobname}

\title{dyngen: a multi-modal simulator for spearheading new single-cell
omics analyses}
\author{Robrecht Cannoodt\textsuperscript{$\dagger{}$,1,2,3} \and Wouter
Saelens\textsuperscript{$\dagger{}$,1,2} \and Louise
Deconinck\textsuperscript{1,2} \and Yvan Saeys\textsuperscript{1,2,*}}
\date{12 May 2021}

@@

\begin{eqnarray*}
  \text{regeffect}_G & = \frac{\text{proptrans}_G(S) - \text{proptrans}_G(S[\text{z}_{T} \leftarrow 0])}{\text{xpr}_{G}}
\end{eqnarray*}

@@

@@

\begin{filecontents}{\jobname.bib}
@Article{aibar_scenicsinglecellregulatory_2017,
	title = {{{SCENIC}}: Single-Cell Regulatory Network Inference and Clustering},
	author = {Sara Aibar and Carmen Bravo Gonz{\a'a}lez-Blas and Thomas Moerman and V{\^a}n Anh Huynh-Thu and Hana Imrichova and Gert Hulselmans and Florian Rambow and Jean-Christophe Marine and Pierre Geurts and Jan Aerts and Joost {van den Oord} and Zeynep Kalender Atak and Jasper Wouters and Stein Aerts},
	year = {2017},
	month = {oct},
	issn = {1548-7091},
	doi = {10.1038/nmeth.4463},
	journal = {Nature Methods},
}

@Article{alpert_alignmentsinglecelltrajectories_2018,
	title = {Alignment of Single-Cell Trajectories to Compare Cellular Expression Dynamics},
	author = {Ayelet Alpert and Lindsay S. Moore and Tania Dubovik and Shai S. Shen-Orr},
	year = {2018},
	month = {apr},
	volume = {15},
	pages = {267--270},
	issn = {1548-7105},
	doi = {10.1038/nmeth.4628},
	abstract = {cellAlign enables quantitative comparisons of expression dynamics within and between single-cell trajectories based on single-cell RNA-seq or mass cytometry data.},
	copyright = {2018 Nature Publishing Group, a division of Macmillan Publishers Limited. All Rights Reserved.},
	file = {/home/rcannood/Zotero/storage/QB7FH2T9/nmeth.html},
	journal = {Nature Methods},
	language = {en},
	number = {4},
}

@Article{bergen_generalizingrnavelocity_2020,
	title = {Generalizing {{RNA}} Velocity to Transient Cell States through Dynamical Modeling},
	author = {Volker Bergen and Marius Lange and Stefan Peidli and F. Alexander Wolf and Fabian J. Theis},
	year = {2020},
	month = {dec},
	volume = {38},
	pages = {1408--1414},
	issn = {1546-1696},
	doi = {10.1038/s41587-020-0591-3},
	abstract = {RNA velocity has opened up new ways of studying cellular differentiation in single-cell RNA-sequencing data. It describes the rate of gene expression change for an individual gene at a given time point based on the ratio of its spliced and unspliced messenger RNA (mRNA). However, errors in velocity estimates arise if the central assumptions of a common splicing rate and the observation of the full splicing dynamics with steady-state mRNA levels are violated. Here we present scVelo, a method that overcomes these limitations by solving the full transcriptional dynamics of splicing kinetics using a likelihood-based dynamical model. This generalizes RNA velocity to systems with transient cell states, which are common in development and in response to perturbations. We apply scVelo to disentangling subpopulation kinetics in neurogenesis and pancreatic endocrinogenesis. We infer gene-specific rates of transcription, splicing and degradation, recover each cell's position in the underlying differentiation processes and detect putative driver genes. scVelo will facilitate the study of lineage decisions and gene regulation.},
	journal = {Nature Biotechnology},
	number = {12},
}

@Article{bjornsson_digitaltwinspersonalize_2019,
	title = {Digital Twins to Personalize Medicine},
	author = {Bergthor Bj{\"o}rnsson and Carl Borrebaeck and Nils Elander and Thomas Gasslander and Danuta R. Gawel and Mika Gustafsson and Rebecka J{\"o}rnsten and Eun Jung Lee and Xinxiu Li and Sandra Lilja and David Mart\'inez-Enguita and Andreas Matussek and Per Sandstr{\"o}m and Samuel Sch{\"a}fer and Margaretha Stenmarker and X. F. Sun and Oleg Sysoev and Huan Zhang and Mikael Benson and {on behalf of the Swedish Digital Twin Consortium}},
	year = {2019},
	month = {dec},
	volume = {12},
	pages = {4},
	issn = {1756-994X},
	doi = {10.1186/s13073-019-0701-3},
	abstract = {Personalized medicine requires the integration and processing of vast amounts of data. Here, we propose a solution to this challenge that is based on constructing Digital Twins. These are high-resolution models of individual patients that are computationally treated with thousands of drugs to find the drug that is optimal for the patient.},
	journal = {Genome Medicine},
	number = {1},
}

@Article{cacchiarelli_aligningsinglecelldevelopmental_2018,
	title = {Aligning {{Single}}-{{Cell Developmental}} and {{Reprogramming Trajectories Identifies Molecular Determinants}} of {{Myogenic Reprogramming Outcome}}},
	author = {Davide Cacchiarelli and Xiaojie Qiu and Sanjay Srivatsan and Anna Manfredi and Michael Ziller and Eliah Overbey and Antonio Grimaldi and Jonna Grimsby and Prapti Pokharel and Kenneth J. Livak and Shuqiang Li and Alexander Meissner and Tarjei S. Mikkelsen and John L. Rinn and Cole Trapnell},
	year = {2018},
	month = {sep},
	volume = {7},
	pages = {258-268.e3},
	issn = {2405-4712},
	doi = {10.1016/j.cels.2018.07.006},
	abstract = {{$<$}h2{$>$}Summary{$<$}/h2{$><$}p{$>$}Cellular reprogramming through manipulation of defined factors holds great promise for large-scale production of cell types needed for use in therapy and for revealing principles of gene regulation. However, most reprogramming systems are inefficient, converting only a fraction of cells to the desired state. Here, we analyze MYOD-mediated reprogramming of human fibroblasts to myotubes, a well-characterized model system for direct conversion by defined factors, at pseudotemporal resolution using single-cell RNA-seq. To expose barriers to efficient conversion, we introduce a novel analytic technique, trajectory alignment, which enables quantitative comparison of gene expression kinetics across two biological processes. Reprogrammed cells navigate a trajectory with branch points that correspond to two alternative decision points, with cells that select incorrect branches terminating at aberrant or incomplete reprogramming outcomes. Analysis of these branch points revealed insulin and BMP signaling as crucial molecular determinants of reprogramming. Single-cell trajectory alignment enables rigorous quantitative comparisons between biological trajectories found in diverse processes in development, reprogramming, and other contexts.{$<$}/p{$>$}},
	file = {/home/rcannood/Zotero/storage/YQH8GP2I/Cacchiarelli et al. - 2018 - Aligning Single-Cell Developmental and Reprogrammi.pdf;/home/rcannood/Zotero/storage/Q5MUK95E/S2405-4712(18)30314-4.html},
	journal = {Cell Systems},
	language = {English},
	number = {3},
	pmid = {30195438},
}

@Article{cannoodt_singlecellomicsdatasets_2018,
	title = {Single-Cell -Omics Datasets Containing a Trajectory},
	author = {Robrecht Cannoodt and Wouter Saelens and Helena Todorov and Yvan Saeys},
	year = {2018},
	month = {oct},
	doi = {10.5281/zenodo.1211532},
	journal = {Zenodo},
}

@TechReport{conover_multiplecomparisonsprocedures_1979,
	title = {On Multiple-Comparisons Procedures},
	author = {William Jay Conover and Ronald L Iman},
	year = {1979},
	institution = {{Technical report, Los Alamos Scientific Laboratory}},
}

@Article{duo_systematicperformanceevaluation_2018,
	title = {A Systematic Performance Evaluation of Clustering Methods for Single-Cell {{RNA}}-Seq Data},
	author = {Angelo Du{\a`o} and Mark D. Robinson and Charlotte Soneson},
	year = {2018},
	volume = {7},
	pages = {1141},
	issn = {2046-1402},
	doi = {10.12688/f1000research.15666.2},
	abstract = {Subpopulation identification, usually via some form of unsupervised clustering, is a fundamental step in the analysis of many single-cell RNA-seq data sets. This has motivated the development and application of a broad range of clustering methods, based on various underlying algorithms. Here, we provide a systematic and extensible performance evaluation of 14 clustering algorithms implemented in R, including both methods developed explicitly for scRNA-seq data and more general-purpose methods. The methods were evaluated using nine publicly available scRNA-seq data sets as well as three simulations with varying degree of cluster separability. The same feature selection approaches were used for all methods, allowing us to focus on the investigation of the performance of the clustering algorithms themselves. We evaluated the ability of recovering known subpopulations, the stability and the run time and scalability of the methods. Additionally, we investigated whether the performance could be improved by generating consensus partitions from multiple individual clustering methods. We found substantial differences in the performance, run time and stability between the methods, with SC3 and Seurat showing the most favorable results. Additionally, we found that consensus clustering typically did not improve the performance compared to the best of the combined methods, but that several of the top-performing methods already perform some type of consensus clustering. All the code used for the evaluation is available on GitHub ( https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison). In addition, an R package providing access to data and clustering results, thereby facilitating inclusion of new methods and data sets, is available from Bioconductor ( https://bioconductor.org/packages/DuoClustering2018).},
	file = {/home/rcannood/Zotero/storage/ASWVFH87/Duò et al. - 2018 - A systematic performance evaluation of clustering .pdf},
	journal = {F1000Research},
	keywords = {Algorithms,Benchmarking,Cluster Analysis,Clustering,Clustering methods,Gene Expression Profiling,Humans,RNA-seq,RNA; Small Cytoplasmic,Sequence Analysis; RNA,Single-Cell Analysis,Single-cell RNA-seq,Software},
	language = {eng},
	pmcid = {PMC6134335},
	pmid = {30271584},
}

@Article{ewing_combiningtumorgenome_2015,
	title = {Combining Tumor Genome Simulation with Crowdsourcing to Benchmark Somatic Single-Nucleotide-Variant Detection},
	author = {Adam D. Ewing and Kathleen E. Houlahan and Yin Hu and Kyle Ellrott and Cristian Caloian and Takafumi N. Yamaguchi and J. Christopher Bare and Christine P'ng and Daryl Waggott and Veronica Y. Sabelnykova and {ICGC-TCGA DREAM Somatic Mutation Calling Challenge Participants} and Liu Xi and Ninad Dewal and Yu Fan and Wenyi Wang and David Wheeler and Andreas Wilm and Grace Hui Ting and Chenhao Li and Denis Bertrand and Niranjan Nagarajan and Qing-Rong Chen and Chih-Hao Hsu and Ying Hu and Chunhua Yan and Warren Kibbe and Daoud Meerzaman and Kristian Cibulskis and Mara Rosenberg and Louis Bergelson and Adam Kiezun and Amie Radenbaugh and Anne-Sophie Sertier and Anthony Ferrari and Laurie Tonton and Kunal Bhutani and Nancy F. Hansen and Difei Wang and Lei Song and Zhongwu Lai and Yang Liao and Wei Shi and Jos{\a'e} Carbonell-Caballero and Joaqu\'in Dopazo and Cheryl C. K. Lau and Justin Guinney and Michael R. Kellen and Thea C. Norman and David Haussler and Stephen H. Friend and Gustavo Stolovitzky and Adam A. Margolin and Joshua M. Stuart and Paul C. Boutros},
	year = {2015},
	month = {jul},
	volume = {12},
	pages = {623--630},
	issn = {1548-7105},
	doi = {10.1038/nmeth.3407},
	abstract = {The detection of somatic mutations from cancer genome sequences is key to understanding the genetic basis of disease progression, patient survival and response to therapy. Benchmarking is needed for tool assessment and improvement but is complicated by a lack of gold standards, by extensive resource requirements and by difficulties in sharing personal genomic information. To resolve these issues, we launched the ICGC-TCGA DREAM Somatic Mutation Calling Challenge, a crowdsourced benchmark of somatic mutation detection algorithms. Here we report the BAMSurgeon tool for simulating cancer genomes and the results of 248 analyses of three in silico tumors created with it. Different algorithms exhibit characteristic error profiles, and, intriguingly, false positives show a trinucleotide profile very similar to one found in human tumors. Although the three simulated tumors differ in sequence contamination (deviation from normal cell sequence) and in subclonality, an ensemble of pipelines outperforms the best individual pipeline in all cases. BAMSurgeon is available at https://github.com/adamewing/bamsurgeon/.},
	copyright = {2015 Nature Publishing Group},
	file = {/home/rcannood/Zotero/storage/NBPKWHK3/Ewing et al. - 2015 - Combining tumor genome simulation with crowdsourci.pdf;/home/rcannood/Zotero/storage/TDZU2TL9/nmeth.html},
	journal = {Nature Methods},
	language = {en},
	number = {7},
}

@Article{ferrell_bistabilitybifurcationswaddington_2012,
	title = {Bistability, {{Bifurcations}}, and {{Waddington}}'s {{Epigenetic Landscape}}},
	author = {James E Ferrell},
	year = {2012},
	month = {jun},
	volume = {22},
	pages = {R458--R466},
	issn = {0960-9822},
	doi = {10.1016/j.cub.2012.03.045},
	abstract = {Waddington's epigenetic landscape is probably the most famous and most powerful metaphor in developmental biology. Cells, represented by balls, roll downhill through a landscape of bifurcating valleys. Each new valley represents a possible cell fate and the ridges between the valleys maintain the cell fate once it has been chosen. Here I examine models of two important developmental processes \$\textbackslash backslash\$textemdash\{\} cell-fate induction and lateral inhibition \$\textbackslash backslash\$textemdash\{\} and ask whether the landscapes for these models at least qualitatively resemble Waddington's picture. For cell-fate induction, the answer is no. The commitment of a cell to a new fate corresponds to the disappearance of a valley from the landscape, not the splitting of one valley into two, and it occurs through a type of bifurcation \$\textbackslash backslash\$textemdash\{\} a saddle-node bifurcation \$\textbackslash backslash\$textemdash\{\} that possesses an intrinsic irreversibility that is missing from Waddington's picture. Lateral inhibition, a symmetrical cell\$\textbackslash backslash\$textendash\{\}cell competition process, corresponds better to Waddington's picture, with one valley reversibly splitting into two through a pitchfork bifurcation. I propose an alternative epigenetic landscape that has numerous valleys and ridges right from the start, with the process of cell-fate commitment corresponding to the irreversible disappearance of some of these valleys and ridges, via cell-fate induction, complemented by the creation of new valleys and ridges through processes like cell\$\textbackslash backslash\$textendash\{\}cell competition.},
	file = {/home/rcannood/Zotero/storage/GK4TJ5BV/Ferrell - 2012 - Bistability, Bifurcations, and Waddington's Epigen.pdf;/home/rcannood/Zotero/storage/HUSMJA22/Ferrell - 2012 - Bistability, Bifurcations, and Waddington's Epigen.pdf;/home/rcannood/Zotero/storage/8VLPBTG2/S0960-9822(12)00332-6.html;/home/rcannood/Zotero/storage/XBTP42Y6/S0960982212003326.html},
	journal = {Current Biology},
	number = {11},
}

@Article{gibson_probabilisticmodelprokaryotic_2000,
	title = {A Probabilistic Model of a Prokaryotic Gene and Its Regulation},
	author = {Michael A. Gibson and Jehoshua Bruck},
	year = {2000},
	journal = {Computational Methods in Molecular Biology: From Genotype to Phenotype, MIT press, Cambridge},
}

@Article{gillespie_exactstochasticsimulation_1977,
	title = {Exact Stochastic Simulation of Coupled Chemical Reactions},
	author = {Daniel T. Gillespie},
	year = {1977},
	month = {dec},
	volume = {81},
	pages = {2340--2361},
	issn = {0022-3654},
	doi = {10.1021/j100540a008},
	file = {/home/rcannood/Zotero/storage/ELV9AQRF/j100540a008.html},
	journal = {The Journal of Physical Chemistry},
	number = {25},
}

@Article{giorgino_computingvisualizingdynamic_2009,
	title = {Computing and {{Visualizing Dynamic Time Warping Alignments}} in {{R}}: {{The}} Dtw {{Package}}},
	author = {Toni Giorgino},
	year = {2009},
	month = {sep},
	doi = {10.18637/jss.v031.i07},
	abstract = {Dynamic time warping is a popular technique for comparing time series, providing both a distance measure that is insensitive to local compression and stretches and the warping which optimally deforms one of the two input series onto the other. A variety of algorithms and constraints have been discussed in the literature. The dtw package provides an unification of them; it allows R users to compute time series alignments mixing freely a variety of continuity constraints, restriction windows, endpoints, local distance definitions, and so on. The package also provides functions for visualizing alignments and constraints using several classic diagram types.},
	journal = {Journal of Statistical Software},
	number = {7},
}

@Article{glass_passingmessagesbiological_2013,
	title = {Passing {{Messages}} between {{Biological Networks}} to {{Refine Predicted Interactions}}},
	author = {Kimberly Glass and Curtis Huttenhower and John Quackenbush and Guo-Cheng Yuan},
	year = {2013},
	month = {may},
	volume = {8},
	pages = {e64832},
	doi = {10.1371/journal.pone.0064832},
	abstract = {Regulatory network reconstruction is a fundamental problem in computational biology. There are significant limitations to such reconstruction using individual datasets, and increasingly people attempt to construct networks using multiple, independent datasets obtained from complementary sources, but methods for this integration are lacking. We developed PANDA (Passing Attributes between Networks for Data Assimilation), a message-passing model using multiple sources of information to predict regulatory relationships, and used it to integrate protein-protein interaction, gene expression, and sequence motif data to reconstruct genome-wide, condition-specific regulatory networks in yeast as a model. The resulting networks were not only more accurate than those produced using individual data sets and other existing methods, but they also captured information regarding specific biological mechanisms and pathways that were missed using other methodologies. PANDA is scalable to higher eukaryotes, applicable to specific tissue or cell type data and conceptually generalizable to include a variety of regulatory, interaction, expression, and other genome-scale data. An implementation of the PANDA algorithm is available at www.sourceforge.net/projects/panda-net.},
	journal = {PLOS ONE},
	number = {5},
}

@Article{graf_forcingcellschange_2009,
	title = {Forcing {{Cells}} to {{Change Lineages}}},
	author = {Thomas Graf and Tariq Enver},
	year = {2009},
	month = {dec},
	volume = {462},
	pages = {587},
	issn = {1476-4687},
	doi = {10.1038/nature08533},
	abstract = {\$\textbackslash textless\$p\$\textbackslash textgreater\$Though it is important for the success of multicellular organisms that their many different cell types remain stable once specified, advances in the production of induced pluripotent stem (iPS) cells from somatic cells by exposure to a transcription factor cocktail emphasize the potential of almost any cell to be reprogrammed. This raises the question of whether interconversion between cell types occurs physiologically or in the context of disease. Thomas Graf and Tariq Enver use examples from the blood cell system to illustrate the principles of transcription factor transdifferentiation, which they conclude is a process fundamentally similar to differentiation and different from iPS reprogramming. They postulate that conversions of one cell type to another can occur as a normal developmental process, and speculate on the prospects of forced lineage reprogramming in regenerative medicine.\$\textbackslash textless\$/p\$\textbackslash textgreater\$},
	journal = {Nature},
	number = {7273},
}

@Article{hache_gengesystematicgeneration_2009,
	title = {{{GeNGe}}: Systematic Generation of Gene Regulatory Networks},
	shorttitle = {{{GeNGe}}},
	author = {Hendrik Hache and Christoph Wierling and Hans Lehrach and Ralf Herwig},
	year = {2009},
	month = {may},
	volume = {25},
	pages = {1205--1207},
	issn = {1367-4803},
	doi = {10.1093/bioinformatics/btp115},
	abstract = {Summary: The analysis of gene regulatory networks (GRNs) is a central goal of bioinformatics highly accelerated by the advent of new experimental techniques, such as RNA interference. A battery of reverse engineering methods has been developed in recent years to reconstruct the underlying GRNs from these and other experimental data. However, the performance of the individual methods is poorly understood and validation of algorithmic performances is still missing to a large extent. To enable such systematic validation, we have developed the web application GeNGe (GEne Network GEnerator), a controlled framework for the automatic generation of GRNs. The theoretical model for a GRN is a non-linear differential equation system. Networks can be user-defined or constructed in a modular way with the option to introduce global and local network perturbations. Resulting data can be used, e.g. as benchmark data for evaluating GRN reconstruction methods or for predicting effects of perturbations as theoretical counterparts of biological experiments.Availability: Available online at http://genge.molgen.mpg.deContact:hache@molgen.mpg.deSupplementary information:Supplementary data are available at Bioinformatics online.},
	journal = {Bioinformatics},
	number = {9},
}

@Article{holm_simplesequentiallyrejective_1979,
	title = {A Simple Sequentially Rejective Multiple Test Procedure},
	author = {Sture Holm},
	year = {1979},
	pages = {65--70},
	journal = {Scandinavian journal of statistics},
}

@Article{huynh-thu_inferringregulatorynetworks_2010,
	title = {Inferring {{Regulatory Networks}} from {{Expression Data Using Tree}}-{{Based Methods}}},
	author = {V{\^a}n Anh Huynh-Thu and Irrthum Alexandre and Wehenkel Louis and Geurts Pierre and Alexandre Irrthum and Louis Wehenkel and Pierre Geurts},
	year = {2010},
	month = {jan},
	volume = {5},
	pages = {e12776},
	issn = {1932-6203},
	doi = {10.1371/journal.pone.0012776},
	abstract = {One of the pressing open problems of computational systems biology is the elucidation of the topology of genetic regulatory networks (GRNs) using high throughput genomic data, in particular microarray gene expression data. The Dialogue for Reverse Engineering Assessments and Methods (DREAM) challenge aims to evaluate the success of GRN inference algorithms on benchmarks of simulated data. In this article, we present GENIE3, a new algorithm for the inference of GRNs that was best performer in the DREAM4 In Silico Multifactorial challenge. GENIE3 decomposes the prediction of a regulatory network between p genes into p different regression problems. In each of the regression problems, the expression pattern of one of the genes (target gene) is predicted from the expression patterns of all the other genes (input genes), using tree-based ensemble methods Random Forests or Extra-Trees. The importance of an input gene in the prediction of the target gene expression pattern is taken as an indication of a putative regulatory link. Putative regulatory links are then aggregated over all genes to provide a ranking of interactions from which the whole network is reconstructed. In addition to performing well on the DREAM4 In Silico Multifactorial challenge simulated data, we show that GENIE3 compares favorably with existing algorithms to decipher the genetic regulatory network of Escherichia coli. It doesn't make any assumption about the nature of gene regulation, can deal with combinatorial and non-linear interactions, produces directed GRNs, and is fast and scalable. In conclusion, we propose a new algorithm for GRN inference that performs well on both synthetic and real gene expression data. The algorithm, based on feature selection with tree-based ensemble methods, is simple and generic, making it adaptable to other types of genomic data and interactions.},
	file = {/home/rcannood/Zotero/storage/3I6PLCQL/Huynh-Thu et al. - 2010 - Inferring Regulatory Networks from Expression Data Using Tree-Based Methods.pdf},
	journal = {PLoS ONE},
	keywords = {Algorithms,Computational Biology,Computational Biology: methods,Escherichia coli,Escherichia coli: genetics,Gene Expression Regulation,Gene Regulatory Networks,Oligonucleotide Array Sequence Analysis},
	number = {9},
	pmid = {20927193},
}

@Article{kanton_organoidsinglecellgenomic_2019,
	title = {Organoid Single-Cell Genomic Atlas Uncovers Human-Specific Features of Brain Development},
	author = {Sabina Kanton and Michael James Boyle and Zhisong He and Malgorzata Santel and Anne Weigert and F{\a'a}tima Sanch\'is-Calleja and Patricia Guijarro and Leila Sidow and Jonas Simon Fleck and Dingding Han and Zhengzong Qian and Michael Heide and Wieland B. Huttner and Philipp Khaitovich and Svante P{\"a}{{\"a}}bo and Barbara Treutlein and J. Gray Camp},
	year = {2019},
	month = {oct},
	volume = {574},
	pages = {418--422},
	issn = {1476-4687},
	doi = {10.1038/s41586-019-1654-9},
	abstract = {Species comparisons using single-cell transcriptomics and accessible chromatin profiling in stem cell-derived cerebral organoids are used to map dynamic gene-regulatory changes that are unique to humans.},
	copyright = {2019 The Author(s), under exclusive licence to Springer Nature Limited},
	file = {/home/rcannood/Zotero/storage/AVPYT7HY/s41586-019-1654-9.html},
	journal = {Nature},
	language = {en},
	number = {7778},
}

@Article{kuijjer_estimatingsamplespecificregulatory_2015,
	title = {Estimating Sample-Specific Regulatory Networks},
	author = {Marieke Lydia Kuijjer and Matthew Tung and GuoCheng Yuan and John Quackenbush and Kimberly Glass},
	year = {2015},
	pages = {1--19},
	abstract = {Biological systems are driven by intricate interactions among the complex array of molecules that comprise the cell. Many methods have been developed to reconstruct network models that attempt to capture those interactions. These methods often draw on large numbers of measured expression samples to tease out subtle signals and infer connections between genes (or gene products). The result is an aggregate network model representing a single estimate for edge likelihoods. While informative, aggregate models fail to capture the heterogeneity that is often represented in a population. Here we propose a method to reverse engineer sample-specific networks from aggregate network models. We demonstrate the accuracy and applicability of our approach in several datasets, including simulated data, microarray expression data from synchronized yeast cells, and RNA-seq data collected from human subjects. We show that these sample-specific networks can be used to study the evolution of network topology across time and to characterize shifts in gene regulation that may not be apparent in the expression data. We believe the ability to generate sample-specific networks will revolutionize the field of network biology and has the potential to usher in an era of precision network medicine.},
	file = {/home/rcannood/Zotero/storage/YECTK66M/Kuijjer et al. - 2015 - Estimating sample-specific regulatory networks.pdf},
}

@Article{kuijjer_estimatingsamplespecificregulatory_2019,
	title = {Estimating {{Sample}}-{{Specific Regulatory Networks}}},
	author = {Marieke Lydia Kuijjer and Matthew George Tung and GuoCheng Yuan and John Quackenbush and Kimberly Glass},
	year = {2019},
	month = {mar},
	volume = {14},
	pages = {226--240},
	issn = {2589-0042},
	doi = {10.1016/j.isci.2019.03.021},
	abstract = {Biological systems are driven by intricate interactions among molecules. Many methods have been developed that draw on large numbers of expression samples to infer connections between genes (or their products). The result is an aggregate network representing a single estimate for the likelihood of each interaction, or ``edge,'' in the network. Although informative, aggregate models fail to capture population heterogeneity. Here we propose a method to reverse engineer sample-specific networks from aggregate networks. We demonstrate our approach in several contexts, including simulated, yeast microarray, and human lymphoblastoid cell line RNA sequencing data. We use these sample-specific networks to study changes in network topology across time and to characterize shifts in gene regulation that were not apparent in the expression data. We believe that generating sample-specific networks will greatly facilitate the application of network methods to large, complex, and heterogeneous multi-omic datasets, supporting the emerging field of precision network medicine.,                                        \textbullet{}               We developed LIONESS to extract single-sample networks from aggregate models                                         \textbullet{}               We tested LIONESS using in silico, yeast, and human expression data                                         \textbullet{}               LIONESS-estimated networks are reproducible, accurate, and biologically meaningful                                         \textbullet{}               Single-sample network analysis highlights important biological processes                                 , Biological Sciences; Bioinformatics; Complex Systems},
	file = {/home/rcannood/Zotero/storage/ITHUQ7NH/Kuijjer et al. - 2019 - Estimating Sample-Specific Regulatory Networks.pdf},
	journal = {iScience},
	pmcid = {PMC6463816},
	pmid = {30981959},
}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  table,
  10pt,
  a4paper]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={dyngen: a multi-modal simulator for spearheading new single-cell omics analyses},
  pdfauthor={Robrecht Cannoodt,1,2,3; Wouter Saelens,1,2,4; Louise Deconinck1,2; Yvan Saeys1,2,},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{tcolorbox}
\usepackage{colortbl}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{fontspec}
\usepackage{pifont}
\usepackage{float}
\usepackage{caption}
\captionsetup[table]{position=bottom}

\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[sorting=none,url=false]{biblatex}
\addbibresource{library.bib}

\title{dyngen: a multi-modal simulator for spearheading new single-cell
omics analyses}
\author{Robrecht Cannoodt\textsuperscript{$\dagger{}$,1,2,3} \and Wouter
Saelens\textsuperscript{$\dagger{}$,1,2,4} \and Louise
Deconinck\textsuperscript{1,2} \and Yvan Saeys\textsuperscript{1,2,*}}
\date{17 May 2021}

@@

\begin{eqnarray}
  \text{regeffect}_G & = \frac{\text{proptrans}_G(S) - \text{proptrans}_G(S[\text{z}_{T} \leftarrow 0])}{\text{xpr}_{G}}
\end{eqnarray}

@@

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

This project has been made possible in part by grant number 2020-218899
from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon
Valley Community Foundation.

@@
